---
title: "RSV_analysis"
author: "Nitsueh Kebere and Jessica McClintock"
date: "5/10/2022"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
setwd("/rprojectnb2/infant-microbiome/work/jessmcc/rsv")
library(tidyverse)
library(devtools) 
library(phyloseq) 
library(HMP) 
library(vegan)
library(picante) 

library(DESeq2)
library(microbiome) 
library(ALDEx2) 
library(metagenomeSeq)
library(dendextend)  
library(selbal) 
library(rms)
library(breakaway)
library(png)
```

### Data Statistics and Visualization 
Possibly come form PathoScope per Brie

```{r message=FALSE, warning=FALSE}
img1 <- readPNG("stat.png")
grid::grid.raster(img1)
```


```{r message=FALSE, warning=FALSE}
img2 <- readPNG("state.png")
grid::grid.raster(img2)
```


```{r message=FALSE, warning=FALSE}
img3 <- readPNG("hiv.png")
grid::grid.raster(img3)
```


```{r message=FALSE, warning=FALSE}
img4 <- readPNG("dth_age.png")
grid::grid.raster(img4)
```


```{r message=FALSE, warning=FALSE}
img5 <- readPNG("gender.png")
grid::grid.raster(img5)
```


```{r message=FALSE, warning=FALSE}
img6 <- readPNG("location.png")
grid::grid.raster(img6)
```


```{r message=FALSE, warning=FALSE}
img7 <- readPNG("duration.png")
grid::grid.raster(img7)
```


### Creating TidyData


```{r message=FALSE, warning=FALSE}
otu_mat <- read_csv("Assay_count_genus.csv")
tax_mat <- read_csv("Assay_taxonomy_genus.csv")
samples_df <- read_csv("Assay_annotation.csv")

# otuTidy <- otu_mat %>%
#   pivot_longer(-X1, names_to = "SampleID", values_to = "Abundance") %>%
#   rename(X1 = "Taxa") 
# 
# samplesTidy <- samples_df %>%
#   rename(X1 = "SampleID")
# 
# zprimeTidy <- inner_join(samplesTidy, otuTidy, by = "SampleID")
# zprimeTidy <- zprimeTidy %>%
#   group_by(subject) %>%
#   mutate(relAbund = Abundance/sum(Abundance))

```

### Create Phyloseq Object
```{r message=FALSE, warning=FALSE}
OTU <- otu_mat %>%
  tibble::column_to_rownames("X1") %>% 
  as.matrix() %>%
  otu_table(taxa_are_rows = TRUE)

TAX <- tax_mat %>% 
  tibble::column_to_rownames("otu") %>% 
  as.matrix() %>%
  tax_table()

samples <- samples_df %>% 
  tibble::column_to_rownames("X1") %>%
  sample_data()

zprime <- phyloseq(OTU, TAX, samples)

```


### Visualizing Relative Abundance 

```{r message=FALSE, warning=FALSE}
relativeAbundance <- function(x){
  x / sum(x)
}

zp_rel_abund <- transform_sample_counts(zprime, relativeAbundance)

plot_bar(zp_rel_abund, fill = "genus") +
  geom_bar(aes(color = genus, fill = genus), stat = "identity", position = "stack") +
  labs(x = "Sample", y = "Relative Abundance\n") +
  facet_wrap(~state, scales = "free") +
  theme(panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```
### Visulaize Relative Abundance with Genera <3.5% pooled to "Other"

```{r message=FALSE, warning=FALSE, fig.width=10}
# otuTidy <- OTU %>%
#   as.data.frame()
# otuTidy <- lapply(otuTidy[,-1],function(x) {x/sum(x)}) %>%
#   as.data.frame()
#   
# row.names(otuTidy) <- otu_mat$X1
# otuTidy <- otuTidy %>% 
#   rownames_to_column("Taxa") %>%
#   pivot_longer(!Taxa, names_to = "Sample", values_to = "relAbund")
# 
# #otuTidy %>% group_by(Taxa) %>% summarise(max = max(relAbund)) %>% arrange(desc(max))
# 
# taxonPooled <- otuTidy %>% 
#   group_by(Taxa) %>% 
#   summarise(pool = mean(relAbund) < 0.05, .groups = "drop")


taxasToMerge <- c()

for (taxa in 1:nrow(otu_table(zp_rel_abund))){
  if (mean(otu_table(zp_rel_abund)[taxa]) < 0.035){
    taxasToMerge <- append(taxasToMerge, row.names(otu_table(zp_rel_abund))[taxa])
  }
}

zpMerged <- merge_taxa(zp_rel_abund, taxasToMerge)

plot_bar(zpMerged, fill = "genus") +
  geom_bar(aes(color = genus, fill = genus), stat = "identity", position = "stack") +
  labs(x = "Sample", y = "Relative Abundance\n") +
  facet_wrap(~state, scales = "free") +
  theme(panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
#N/A in graph represent all merged taxa

#Separated by Genus to see individual fluctuations  
plot_bar(zpMerged, fill = "genus") +
  geom_bar(aes(color = genus, fill = genus), stat = "identity", position = "stack") +
  labs(x = "Sample", y = "Relative Abundance\n") +
  facet_grid(genus~state, scales = "free") +
  theme(panel.background = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

  
```

### Mean Abundance by Disease State - Genus 
```{r messafe=FALSE, warning=FALSE, fig.width=10}
plot_bar(zpMerged, x = "state", fill = "genus")+ geom_bar(aes(color=genus, fill=genus), stat = "identity", position = "stack")
```

### Visulaizing Abundance Grouped by Genus

```{r message=FALSE, warning=FALSE, fig.width=10}
zp_genus <- tax_glom(zp_rel_abund, "genus") #relative abundance
zp_genus_counts <- tax_glom(zprime, "genus") #abundance in counts

psmelt(zp_genus) %>%
  ggplot(data = ., aes(x = state, y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  geom_jitter(aes(color = genus), height = 0, width = .2) +
  labs(title = "Relative Abundance based on RSV Status", x = "", y = "Relative Abundance\n") +
  facet_wrap(~ OTU, scales = "free") +
  theme(legend.position = "none")


```
### Testing for a difference in genus level abundance 

```{r, message=FALSE, warning=FALSE}
zp_pos <- subset_samples(zp_genus_counts, state == "pos")
zp_neg <- subset_samples(zp_genus_counts, state =="neg")

zp_pos_otu <- otu_table(zp_pos) %>%
  data.frame() %>%
  t() %>%
  as.data.frame() %>%
  mutate(Other = Dolosigranulum + Gemella + Porphyromonas + Serratia + Veillonella) %>%
  dplyr::select(-Dolosigranulum, -Gemella, -Porphyromonas, -Serratia, -Veillonella) %>%
  as.matrix()

zp_neg_otu <- otu_table(zp_neg) %>%
  data.frame() %>%
  t() %>%
  as.data.frame() %>%
  mutate(Other = Dolosigranulum + Gemella + Porphyromonas + Serratia + Veillonella) %>%
  dplyr::select(-Dolosigranulum, -Gemella, -Porphyromonas, -Serratia, -Veillonella) %>%
  as.matrix()

group_data <- list(zp_pos_otu, zp_neg_otu)
(xdc <- HMP::Xdc.sevsample(group_data)) 
```

### Visualizing Abundance Grouped by Location

```{r message=FALSE, warning=FALSE, fig.width=10}

psmelt(zp_genus) %>%
  ggplot(data = ., aes(x = location, y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  geom_jitter(aes(color = genus), height = 0, width = .2) +
  labs(title = "Relative Abundance Based on Location of Death", x = "", y = "Abundance\n") +
  facet_wrap(~ OTU, scales = "free") +
  theme(legend.position = "none", plot.title=element_text(hjust=0.5, size=18))

```
### Testing for a difference in genus level abundance by location 

```{r, message=FALSE, warning=FALSE}
zp_BID <- subset_samples(zp_genus_counts, location == "BID")
zp_Hospital <- subset_samples(zp_genus_counts, location == "Hospital")

zp_BID_otu <- otu_table(zp_BID) %>%
  data.frame() %>%
  t() %>%
  as.data.frame() %>%
  mutate(Other = Dolosigranulum + Gemella + Porphyromonas + Serratia + Veillonella) %>%
  dplyr::select(-Dolosigranulum, -Gemella, -Porphyromonas, -Serratia, -Veillonella) %>%
  as.matrix()

zp_Hospital_otu <- otu_table(zp_Hospital) %>%
  data.frame() %>%
  t() %>%
  as.data.frame() %>%
  mutate(Other = Dolosigranulum + Gemella + Porphyromonas + Serratia + Veillonella) %>%
  dplyr::select(-Dolosigranulum, -Gemella, -Porphyromonas, -Serratia, -Veillonella) %>%
  as.matrix()

group_data <- list(zp_BID_otu, zp_Hospital_otu)
(xdc_location <- HMP::Xdc.sevsample(group_data)) 
```

### Abundance by HIV status
```{r message=FALSE, warning=FALSE, fig.width=10}

psmelt(zp_genus) %>%
  ggplot(data = ., aes(x = as.factor(mother_hiv), y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  geom_jitter(aes(color = genus), height = 0, width = .2) +
  labs(title = "Relative Abundance Based on Mother's HIV Status", x = "", y = "Abundance\n") +
  facet_wrap(~ OTU, scales = "free") +
  theme(legend.position = "none", plot.title=element_text(hjust=0.5, size=18))

```
### Testing for a difference in genus level abundance by location 

```{r, message=FALSE, warning=FALSE}
zp_HIV_neg <- subset_samples(zp_genus_counts, mother_hiv == 0)
zp_HIV_pos <- subset_samples(zp_genus_counts, mother_hiv == 1)

zp_HIV_neg_otu <- otu_table(zp_HIV_neg) %>%
  data.frame() %>%
  t() %>%
  as.data.frame() %>%
  mutate(Other = Dolosigranulum + Gemella + Porphyromonas + Serratia + Veillonella) %>%
  dplyr::select(-Dolosigranulum, -Gemella, -Porphyromonas, -Serratia, -Veillonella) %>%
  as.matrix()

zp_HIV_pos_otu <- otu_table(zp_HIV_pos) %>%
  data.frame() %>%
  t() %>%
  as.data.frame() %>%
  mutate(Other = Dolosigranulum + Gemella + Porphyromonas + Serratia + Veillonella) %>%
  dplyr::select(-Dolosigranulum, -Gemella, -Porphyromonas, -Serratia, -Veillonella) %>%
  as.matrix()

group_data <- list(zp_HIV_neg_otu, zp_HIV_pos_otu)
(xdc_location <- HMP::Xdc.sevsample(group_data)) 
```

### Abundance based on Length of Hospital Stay
```{r message=FALSE, warning=FALSE, fig.width=10}

psmelt(zp_genus) %>%
  ggplot(data = ., aes(x = hosp_duration, y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  geom_jitter(aes(color = genus), height = 0, width = .2) +
  labs(title = "Relative Abundance Based on Length of Hospital Stay", x = "", y = "Abundance\n") +
  facet_wrap(~ OTU, scales = "free") +
  theme(legend.position = "none", plot.title=element_text(hjust=0.5, size=22))

```
### Abundance based on Gender
```{r message=FALSE, warning=FALSE, fig.width=10}

psmelt(zp_genus) %>%
  #filter(gender != 8) %>%
  ggplot(data = ., aes(x = as.factor(gender), y = Abundance)) +
  geom_boxplot(outlier.shape  = NA) +
  geom_jitter(aes(color = genus), height = 0, width = .2) +
  labs(title = "Relative Abundance based on Gender", x = "Gender", y = "Abundance\n") +
  facet_wrap(~ OTU, scales = "free") +
  theme(legend.position = "none", plot.title=element_text(hjust=0.5, size=22))

```
### Heirarchical Clustering 


```{r, message=FALSE, warning=FALSE, fig.width=15}
#Extract OTU table and compute BC
zp_rel_otu <- otu_table(zp_rel_abund) %>%
  as.data.frame() %>%
  t() #vegan requires each row to be a sample and column to be data

bc_dist <- vegdist(zp_rel_otu, method = "bray")
min(bc_dist)
max(bc_dist)
mean(bc_dist)
median(bc_dist)

ward <- as.dendrogram(hclust(bc_dist, method = "ward.D2"))
#Provide color codes
meta <- data.frame(sample_data(zp_rel_abund))
colorCode <- c(pos = "red", neg = "blue")
labels_colors(ward) <- colorCode[meta$state][order.dendrogram(ward)]
#Plot
plot(ward)

```


### HeatMap 


```{r, message=FALSE, warning=FALSE, fig.width=15}

p <- plot_heatmap(zprime, method = "NMDS", distance = "bray",
             sample.label = "state", sample.order = "state")
axis_color <- ifelse(p$scales$scales[[1]]$labels == "pos", "red", "blue")
p + theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 10, colour = axis_color), plot.margin = unit(c(1,1,1,1),"cm"))


```


### Alpha Diversity 

```{r}
ggplot(data = data.frame("total_reads" =  phyloseq::sample_sums(zprime),
                         "observed" = phyloseq::estimate_richness(zprime, measures = "Observed")[, 1]),
       aes(x = total_reads, y = observed)) +
  geom_point() +
  geom_smooth(method="lm", se = FALSE) +
  labs(x = "\nTotal Reads", y = "Observed Richness\n")

```


```{r, message=FALSE, warning=FALSE}

#(zp_rare <- rarefy_even_depth(zprime, rngseed = 123, replace = FALSE))
# set seed to 123 to create a repeatable random sampling, not recommended to use subsampling

adiv <- data.frame(
  "Observed" = estimate_richness(zprime, measures = "Observed"),
  "Shannon" = estimate_richness(zprime, measures = "Shannon"),
  "Simpson" = estimate_richness(zprime, measures = "Simpson"),
  "Status" = sample_data(zprime)$state)
#head(adiv)

adiv %>%
  gather(key = metric, value = value, c("Observed", "Shannon", "Simpson")) %>%
  mutate(metric = factor(metric, levels = c("Observed", "Shannon", "Simpson"))) %>%
  ggplot(aes(x = Status, y = value)) +
  geom_boxplot(outlier.color = "black") +
  geom_jitter(aes(color = Status), height = 0, width = .2) +
  labs(x = "", y = "") +
  facet_wrap(~ metric, scales = "free") +
  theme(legend.position="none")


```


```{r, message=FALSE, warning=FALSE}
adiv %>%
  group_by(Status) %>%
  summarise(median_observed = median(Observed),
            median_shannon = median(Shannon),
            median_simpson = median(Simpson))

# Test for normality
shapiro.test(subset(adiv, Status == "pos")$Shannon)
shapiro.test(subset(adiv, Status == "neg")$Shannon)
shapiro.test(subset(adiv, Status == "pos")$Simpson)
shapiro.test(subset(adiv, Status == "neg")$Simpson)

# pos samples are not normally distributed, neg samples are, so use 
wilcox.test(Observed ~ Status, data = adiv, exact = FALSE, conf.int = TRUE)
wilcox.test(Shannon ~ Status, data = adiv, conf.int = TRUE) 
wilcox.test(Simpson ~ Status, data = adiv, conf.int = TRUE) 

```


### Beta Diversity 


#### PCA


```{r, message=FALSE, warning=FALSE}
zp_clr <- microbiome::transform(zprime, "clr")  #centered log ratio, log(x1/mean(x))

#PCA via phyloseq
ord_clr <- phyloseq::ordinate(zp_clr, "RDA")
#Plot scree plot
phyloseq::plot_scree(ord_clr) + 
  geom_bar(stat="identity", fill = "blue") +
  labs(x = "\nAxis", y = "Proportion of Variance\n")
#head(ord_clr$CA$eig)  

```

```{r}
#Scale axes and plot ordination
clr1 <- ord_clr$CA$eig[1] / sum(ord_clr$CA$eig)
clr2 <- ord_clr$CA$eig[2] / sum(ord_clr$CA$eig)
phyloseq::plot_ordination(zprime, ord_clr, type="samples", color="state") + 
  geom_point(size = 2) +
  coord_fixed(clr2 / clr1) +
  stat_ellipse(aes(group = state), linetype = 2)
#We see some clustering by RSV status
```

#### PCoA


```{r, message=FALSE, warning=FALSE}
#Generate distance matrix
clr_dist_matrix <- phyloseq::distance(zp_clr, method = "euclidean") 
#ADONIS test
vegan::adonis(clr_dist_matrix ~ phyloseq::sample_data(zp_clr)$state)

dispr <- vegan::betadisper(clr_dist_matrix, phyloseq::sample_data(zp_clr)$state)
#dispr

plot(dispr, main = "Ordination Centroids and Dispersion Labeled: Aitchison Distance", sub = "")

```

```{r}
boxplot(dispr, main = "", xlab = "")
permutest(dispr)

```

#### NMDS


```{r, message=FALSE, warning=FALSE}

zprime.ord <- ordinate(zprime, "NMDS", "bray")

plot_ordination(zprime, zprime.ord, type="samples", color="state", 
                title="Samples") + geom_point(size=3)



```

### Differential Abundance Testing - RSV state

```{r, message=FALSE, warning=FALSE}
#Generate data.frame with OTUs and metadata
zp_wilcox <- data.frame(t(data.frame(phyloseq::otu_table(zp_clr))))
zp_wilcox$state <- phyloseq::sample_data(zp_clr)$state

#Define functions to pass to map
wilcox_model <- function(df){
  wilcox.test(abund ~state, data = df)
}
wilcox_pval <- function(df){
  wilcox.test(abund ~state, data = df)$p.value
}

#Create nested data frames by OTU and loop over each using map 
wilcox_results <- zp_wilcox %>%
  gather(key = Genus, value = abund, -state) %>%
  group_by(Genus) %>%
  nest() %>%
  mutate(wilcox_test = map(data, wilcox_model),
         p_value = map(data, wilcox_pval))       

#Show results
#head(wilcox_results)
#head(wilcox_results$data[[1]])
#wilcox_results$p_value[[1]]


#Unnesting
wilcox_results <- wilcox_results %>%
  dplyr::select(Genus, p_value) %>%
  unnest()


head(wilcox_results)

#Adding taxonomic labels
taxa_info <- data.frame(tax_table(zp_clr))
taxa_info <- taxa_info %>% rownames_to_column(var = "Genus")
#Computing FDR corrected p-values
wilcox_results <- wilcox_results %>%
#   full_join(taxa_info) %>%
   arrange(p_value) %>%
   mutate(BH_FDR = p.adjust(p_value, "BH")) %>%
   #filter(BH_FDR < 0.05) %>%
   dplyr::select(Genus, p_value, BH_FDR, everything())

print.data.frame(wilcox_results) 

```

### Differential Abundance - Location
```{r, message=FALSE, warning=FALSE}
zp_wilcox <- data.frame(t(data.frame(phyloseq::otu_table(zp_clr))))
zp_wilcox$location <- phyloseq::sample_data(zp_clr)$location

#Define functions to pass to map
wilcox_model <- function(df){
  wilcox.test(abund ~location, data = df)
}
wilcox_pval <- function(df){
  wilcox.test(abund ~location, data = df)$p.value
}

#Create nested data frames by OTU and loop over each using map 
wilcox_results <- zp_wilcox %>%
  gather(key = Genus, value = abund, -location) %>%
  group_by(Genus) %>%
  nest() %>%
  mutate(wilcox_test = map(data, wilcox_model),
         p_value = map(data, wilcox_pval))       

#Show results
#head(wilcox_results)
#head(wilcox_results$data[[1]])
#wilcox_results$p_value[[1]]


#Unnesting
wilcox_results <- wilcox_results %>%
  dplyr::select(Genus, p_value) %>%
  unnest()


head(wilcox_results)

#Adding taxonomic labels
taxa_info <- data.frame(tax_table(zp_clr))
taxa_info <- taxa_info %>% rownames_to_column(var = "Genus")
#Computing FDR corrected p-values
wilcox_results <- wilcox_results %>%
#   full_join(taxa_info) %>%
   arrange(p_value) %>%
   mutate(BH_FDR = p.adjust(p_value, "BH")) %>%
   filter(BH_FDR < 0.05) %>%
   dplyr::select(Genus, p_value, BH_FDR, everything())

print.data.frame(wilcox_results) 

```
### Differential Abundance - HIV Exposure
```{r, message=FALSE, warning=FALSE}
zp_wilcox <- data.frame(t(data.frame(phyloseq::otu_table(zp_clr))))
zp_wilcox$mother_hiv <- phyloseq::sample_data(zp_clr)$mother_hiv

#Define functions to pass to map
wilcox_model <- function(df){
  wilcox.test(abund ~mother_hiv, data = df)
}
wilcox_pval <- function(df){
  wilcox.test(abund ~mother_hiv, data = df)$p.value
}

#Create nested data frames by OTU and loop over each using map 
wilcox_results <- zp_wilcox %>%
  gather(key = Genus, value = abund, -mother_hiv) %>%
  group_by(Genus) %>%
  nest() %>%
  mutate(wilcox_test = map(data, wilcox_model),
         p_value = map(data, wilcox_pval))       

#Show results
#head(wilcox_results)
#head(wilcox_results$data[[1]])
#wilcox_results$p_value[[1]]


#Unnesting
wilcox_results <- wilcox_results %>%
  dplyr::select(Genus, p_value) %>%
  unnest()


head(wilcox_results)

#Adding taxonomic labels
taxa_info <- data.frame(tax_table(zp_clr))
taxa_info <- taxa_info %>% rownames_to_column(var = "Genus")
#Computing FDR corrected p-values
wilcox_results <- wilcox_results %>%
#   full_join(taxa_info) %>%
   arrange(p_value) %>%
   mutate(BH_FDR = p.adjust(p_value, "BH")) %>%
   filter(BH_FDR < 0.05) %>%
   dplyr::select(Genus, p_value, BH_FDR, everything())

print.data.frame(wilcox_results) 

```

### Linear Regression
```{r, message=FALSE, warning=FALSE}
gemellaCounts <- as.data.frame(t(as.data.frame(otu_table(zprime)["Gemella"])))
gemellaInfo <- merge(meta, gemellaCounts, by=0)

for (row in 1:nrow(gemellaInfo)){
   if(gemellaInfo[row, "state"] == "pos"){
     gemellaInfo[row, "state"] <- 1
   }else if(gemellaInfo[row, "state"] == "neg"){
     gemellaInfo[row, "state"] <- 0
   }

  if(gemellaInfo[row, "location"] == "BID"){
    gemellaInfo[row, "location"] <- 0
  }else if(gemellaInfo[row, "location"] == "Hospital"){
    gemellaInfo[row, "location"] <- 1
  }
}
  
fit0 <- lm(Gemella ~ 1, gemellaInfo)
fit1 <- lm(Gemella ~ state, gemellaInfo)
fit2 <- lm(Gemella ~ state + mother_hiv, gemellaInfo)
fit3 <- lm(Gemella ~ state + mother_hiv + location, gemellaInfo)

summary(fit0)
summary(fit1)
summary(fit2)
summary(fit3)

anova(fit0, fit1, test='F')

```
